Omics-based Single Step Trait Prediction

### 2017-05-12 paper/manuscript.tex
Update methods

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the $l$-th group of inbred lines and the 
$o$-th predictor ($\mathbf{W}_{lo}$) has dimensions $n_{l} \times p_{o}$ where 
$n_{l}$ pertains to the number of genotypes in the $l$-th group and $p_{o}$
pertains to the number of features.
Note that $l$ is only defined for \textit{Exp1} where a distinction between 
the two heterotic groups is necessary whereas $l$ can be ignored for all 
models pertaining to the inbred lines of the maize diversity panel in 
\textit{Exp2}.
To avoid mentioning special cases, we will use the general notation with 
the index $l$ throughout this section, though.
For the diversity panel maize lines, $n_{l}$ corresponds to the number of 
genotypes whereas, in the case of the hybrid data, $n_{l}$ corresponds to the 
number of parent lines in the $l$-th heterotic group.
In generating kernels for each predictor, all features in $\mathbf{W}_{lo}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

\begin{equation} \label{eq:GenomicRelationship}
  \mathbf{K}_{lo} = \frac{1}{W_{lo}} \mathbf{W}_{lo} \mathbf{W}_{lo}^{\top},
\end{equation}

where $W_{lo}$ denotes the number of features \cite{VanRaden2008}.
In the case of pedigree data (P), coancestry coefficients were used directly
for $\mathbf{K}_{lo}$.

@@

\begin{equation} \label{eq:KBLUPModel}
  \mathbf{y} = \mu + 
  \sum_{l=1}^{L} \mathbf{Z}_{lo} \mathbf{g}_{lo} +
  \mathbf{\epsilon},
\end{equation}

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\epsilon$ is the residual
error.
For the hybrid data from \textit{Exp1}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{lo}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp2}, we fit random 
genotype effects $\mathbf{g}_{lo}$ (\textit{i.e.}, breeding values) of the 
lines for the $o$-th predictor.
For both experiments, genetic effects $\mathbf_{g}_{lo}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{lo}$.
The random effects ($\mathbf{g}_{lo}$) have expectation zero and variance equal 
to $\mathbf{K}_{lo} \sigma^{2}_{{g}^{lo}}$ and $\mathbf{I} \sigma^2_{\epsilon}$ 
for the residual error.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single step prediction}
The breeding values in a prediction model are in general given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{W}\boldsymbol{\hat{\alpha}},
\end{equation}

where $\boldsymbol{\hat{\alpha}}$ is the solution to a ridge regression model,
which is equivalent to the BLUP model.

Consider now the situation in which one predictor is complete in the sense that
it contains information on the whole population whereas a second predictor is
incomplete and covers only a subset of the population.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Further, the covariates in $\mathbf{W}$ are centered.
The vector $\mathbf{g}_1$ can be written as the sum of the conditional 
expectation given $\mathbf{g}_2$ and a residual error term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K_{12}}\mathbf{K_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K_{12}}\mathbf{K_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon}
\end{align}

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} = 
\begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix},
 \mathbf{W} = 
\begin{bmatrix}
  Z_1\hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix},
 \mathbf{U} = 
\begin{bmatrix}
  Z_1 \\
  0 
 \end{bmatrix}
\end{equation}

>>>>

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the $l$-th group of inbred lines 
($\mathbf{W}_{l}$) has dimensions $n_{l} \times p$ where $n_{l}$ pertains to 
the number of genotypes in the $l$-th group and $p$ pertains to the number of 
features.
Note that $l$ is only defined for \textit{Exp1} where a distinction between 
the two heterotic groups is necessary whereas $l$ can be ignored for all 
models pertaining to the inbred lines of the maize diversity panel in 
\textit{Exp2}.
To avoid mentioning special cases, we will use the general notation with 
the index $l$ throughout this section, though.
For the diversity panel maize lines, $n_{l}$ corresponds to the number of 
genotypes whereas, in the case of the hybrid data, $n_{l}$ corresponds to the 
number of parent lines in the $l$-th heterotic group.
In generating kernels for each predictor, all features in $\mathbf{W}_{l}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

\begin{equation} \label{eq:GenomicRelationship}
  \mathbf{K}_{l} = \frac{1}{W_{l}} \mathbf{W}_{l} \mathbf{W}_{l}^{\top},
\end{equation}

where $W_{l}$ denotes the number of features \cite{VanRaden2008}.
In the case of pedigree data (P), coancestry coefficients were used directly
for $\mathbf{K}_{l}$.

@@

\begin{equation} \label{eq:KBLUPModel}
  \mathbf{y} = \mu + 
  \sum_{l=1}^{L} \mathbf{Z}_{l} \mathbf{g}_{l} + \mathbf{e},
\end{equation}

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from \textit{Exp1}, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from \textit{Exp2}, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values) of the 
lines for the $o$-th predictor.
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
"omics" predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single step prediction}
For simplicity, let us assume throughout this section that we have $L=1$.
If the breeding values are fully captured by the features in $\mathbf{W}$,
Eq.~\ref{eq:KBLUPModel} can then be expressed as

\begin{align} \label{eq:KBLUPModelSimplified}
	\mathbf{y} &= \mu + \mathbf{Z} \mathbf{g} + \boldsymbol{e} \\
	&= \mu + \mathbf{Z} \mathbf{W} \boldsymbol{\alpha} + \boldsymbol{e}
\end{align}
\cite{Fernando2014} since the ridge regression model is equivalent to the BLUP
model \cite{Ruppert2003}.

Hence, the breeding values in a prediction model are given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{W}\boldsymbol{\hat{\alpha}},
\end{equation}

where $\boldsymbol{\hat{\alpha}}$ is the solution to a ridge regression model.

Consider now the situation in which one predictor is complete in the sense that
it contains information on the whole population whereas a second predictor is
incomplete and covers only a subset of the population.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Further, the covariates in $\mathbf{W}$ are centered.
In this case we need to partition the vector of breeding values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} breeding values of animals 
covered only by the complete predictor) and a component 
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} breeding 
values of animals covered by both predictors).
The vector $\mathbf{g}_1$ can be written as the sum of the conditional 
expectation given $\mathbf{g}_2$ and a residual error term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K_{12}}\mathbf{K_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K_{12}}\mathbf{K_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\mathbf{K}_{ij}$ are partitions of $\mathbf{K}$ that correspond to 
$\mathbf{g}_{1}$ and $\mathbf{g}_{2}$, respectively \cite{Fernando2014}.

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} = 
\begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix},
 \mathbf{W} = 
\begin{bmatrix}
  Z_1\hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix},
 \mathbf{U} = 
\begin{bmatrix}
  Z_1 \\
  0 
 \end{bmatrix},
\end{equation}

where $\hat{\mathbf{W}}_{1} = \mathbf{K}_{12} \mathbf{K}_{22}^{-1} \mathbf{W}_{2}$.
